---
title: "DunedinPACE: A DNA methylation biomarker of the pace of aging"
author: Vignette created by - Calen P. Ryan, PhD
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  % \VignetteIndexEntry{DunedinPACE: A DNA methylation biomarker of the pace of aging}
  % \VignetteEngine{knitr::rmarkdown}
  % \VignetteEncoding{UTF-8}
---

# Introduction to DunedinPACE

`DunedinPACE` is a novel blood biomarker of the pace of aging for gerontology and geroscience. It shows high test-retest reliability, is associated with morbidity, disability, and mortality, and indicated faster aging in young adults with childhood adversity. DunedinPACE effect-sizes are similar to GrimAge Clock effect-sizes. In analysis of incident morbidity, disability, and mortality, DunedinPACE and added incremental prediction beyond GrimAge.

Citation for the original paper describing `DunedinPACE()` in more detail is provided below:

 * Belsky DW, Caspi A, Corcoran DL, Sugden K, Poulton R, Arseneault L, Baccarelli A, Chamarti K, Gao X, Hannon E, et al. 2022. DunedinPACE, a DNA methylation biomarker of the pace of aging. Deelen J, editor. eLife. 11:e73420. doi:10.7554/eLife.73420.





# The `PACEProjector` function

The `PACEProjector` calculator requires a matrix or data.frame of beta values where rownames are probe ids and column names correspond to sample names. Beta values must be numeric and missing values must be coded as 'NA'. 

```{r}
#Load the library
library(DunedinPACE)
```


```{r}
# Load the data
data("example_betas")
# Extract matrix of beta values
beta_matrix <-SummarizedExperiment::assay(example_betas)

```

```{r}
# Dimensions of beta matrix
dim(beta_matrix)

# Head of beta matrix
# Data should look like this

head(beta_matrix)
```

Generate the `DunedinPACE` projections as follows:

```{r}
pace_values <-PACEProjector(beta_matrix)
```

The output will look like this
```{r}
pace_values
```

Which you can extract as follows:
```{r}
pace_values$DunedinPACE
```

The output, DunedinPACE, is equivalent to the pace of aging (in years) expected over a 1-year period. Higher values are equivalent to faster pace of aging, whereas lower values are equivalent to slower pace of aging.


---

The `PACEProjector` uses data from 20,000 CpGs. This includes 173 probes for PACE projection, and a background list that is used for normalization.

It is possible to account for missing data using the `proportionOfProbesRequired` argument in the `PACEProjector()` function, which by default is set to 0.8. If you run `PACEProjector()` on a full matrix of 450k or EPIC array data, you should not encounter any problems. If you are missing a lot of data, you can optionally lower this threshold to force the calculator to do its job. However, we do not recommend excluding the 19827 probes used for normalization and calculating DunedinPACE using only the 173 DunedinPACE associated probes, as this could affect DunedinPACE estimates. 



# The `getRequiredProbes()` function

CpGs used in the PACE projector can be accessed using the `getRequiredProbes()` command. For a list 173 CpGs used directly for DunedinPACE projection, use the argument `getRequiredProbes(backgroundList = FALSE)`. For a list of all probes used in the `PACEProjector()`, which includes including an additional 19827 probes used for normalization, use the argument `getRequiredProbes(backgroundList = TRUE)`

**Again, we do not recommend excluding the 19827 probes used for normalization and calculating DunedinPACE using only the 173 DunedinPACE associated probes, as this could affect DunedinPACE estimates.  **

```{r}
getRequiredProbes(backgroundList = FALSE)
```





